---
title: "Tutorial"
author: "Bbsharma1"
date: "4/19/2020"
output: html_document
---

```{r}
## Libraries
library(filesstrings) ## to move/copy files in directory
library(lattice) ## for plotting 
library(ggplot2) ## for plotting
```


# Chapter 4 - Exploratory Data Analysis

## Graphs
```{r}
# Download file

fileUrl <- "https://raw.githubusercontent.com/jtleek/modules/master/04_ExploratoryAnalysis/exploratoryGraphs/data/avgpm25.csv"

  if(!file.exists("./Tutorial_Files/EPA.csv") | !file.exists("./Tutorial_Files/archive/EPA.csv"))
  {
    download.file(fileUrl, destfile = "./Tutorial_Files/EPA.csv",  mode = "wb", method = "curl")
    if(!file.exists("./Tutorial_Files/archive")){dir.create("./Tutorial_Files/archive")}
    file.copy("./Tutorial_Files/EPA.csv", "./Tutorial_Files/archive/")  ## backup original file
    EPA_dw_time <- date()
    print("File downloaded")
  } else {
      print("file found, not downloading")
    }
```

```{r}
# Read File

epa_pp <- read.csv("./Tutorial_Files/EPA.csv", colClasses = c("numeric", "character", "factor", "numeric", "numeric"))
epa_pp
## pm25 is the 3 year average (2008 - 2010)
## fips: county identifier
## long/lat is coordinates for the monitoring device

summary(epa_pp)
## initially we can see that the max is 18.441 meaning there is a location where they are way above 12 limit
```

#### plotting our data to view - Histogram
```{r}

hist(epa_pp$pm25, col = "green")

## Observartion: Bulk of data is between 9 and 11

hist(epa_pp$pm25, col = "green", breaks = 100)

## Making the bars smaller and getting better visuals

hist3 <- hist(epa_pp$pm25, col = "green")
abline(v = 12, lwd = 2)
abline(v = median(epa_pp$pm25), col = "magenta", lwd = 4)
hist3
## 'v = ' is vertical line, 'lwd' is line width
## Histograms do not innatly show the median so we entered it outself.


# Multiple Histograms

par(mfrow = c(2,1), mar = c(4,4,2,1))
hist_east <- hist(subset(epa_pp, region == "east")$pm25, col = "green")
hist_west <- hist(subset(epa_pp, region == "west")$pm25, col = "green")
##Western counties are lower in averagem but higher extreme values
```

#### plotting our data to view - Boxplot
```{r}
boxplot(epa_pp$pm25, col = "blue")
abline(h = 12)
## abline sets a line in graph, and we can see there are outliers. Upper quartile is 15, first quartile is at 4


# MultipleBoxplot
boxplot(pm25 ~ region, data = epa_pp, col = "red")
## Observation: pm25 is higher on the east than on the west, however has larger spread (outliers exist at higher rates)
##Observation: We use the R formula y ~ x to show that y  depends on x, so in this case, pm25 depends on region!
```


#### plotting our data to view - Barplot
```{r}
# Best for categorical data 

barplot(table(epa_pp$region), col = "wheat", main = "Number of counties in each reigon")
## Observation: As shown, we get a overview of the different reigons, and number of counties in each reagion
## Observation: The table(epa_pp$riegon) will summarize the data 
```


#### plotting our data to view - Scatterplot
```{r}

plot1 <- with(epa_pp, plot(latitude, pm25))
abline(h = 12, lwd = 2, lty = 2)
## There is no particular trend here, but we can observe that pm25 is around 10-11 when lattitude is between 30 - 50

# Adding Color
plot2 <- with(epa_pp, plot(latitude, pm25, col = region))
abline(h = 12, lwd = 2, lty = 2)
## Observation: scatter plots are now seperated by reigons, east and west through black and red colors, respectively.



#Multiple Scattor plots

# We can do the above through different colors, or we can plot them seperatly like below:

par(mfrow = c(1,2), mar = c(5,4,2,1))
scat_west <- with(subset(epa_pp, region == "west"), plot(latitude, pm25, main = "West"))
scat_east <- with(subset(epa_pp, region == "east"), plot(latitude, pm25, main = "East"))
## Both cases, higher pollution counties tend to be in the middle lattitudes
```



## Plot


### Lattice plot
```{r}
## using data found in lattice plot, lets plot the below: 
state <- data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4,1))
## Observation: indicates we're plotting life expectancy as it depends on income for each region

```


### ggplot2
```{r}
## using mpg data from ggplot package
data(mpg)
qplot(displ, hwy, data = mpg)
## Observation: x asix is displ (engine size) and y axis is hiway milage for that car. We can see that as engine size increases,
## Observation: hway mileage decreases
```



### Base Graphics 
```{r}
# Simple Histogram example

library(datasets) ## load to  play with data
histogram(airquality$Ozone)


# Simple Scattor plot example 
with(airquality, plot(Wind, Ozone))
## Observation: Wind is X axis, Ozone is Y axis)


# Simple boxplot example
airquality <- transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
## Observation: boxplot of distribution by month. X axis is Month, Y axis is ozone

```


#### Base scatter plots, adding annotation example
```{r}

with(airquality, plot(Wind, Ozone))
## Observation: X label: Wind, Ylabel: Ozone. X axis - 0 - 20, y axis: 0 - 150.

title(main = "Ozone and Wind in NY City")
## Observation: Title given. Also, we must rerun the whole segment of code and not just this line.

with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in NY City"))
## Another way for the above, putting title directly into the plot

```

#### Scatterplot Annotation: Adding colors to different parts of the graph
```{r}
# Here, we will subset all of data points for the month of May to output a different color on the plot
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in NY City")) ## draw initial graph
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
```

```{r}
# Adding more variations to graph
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in NY City", type = "n")) ## draw initial graph where n signifies no data to be added
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))

with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
## Observation: Colors the remaining circles to red becuase Month not equal to 5

legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))
## Observation: legend box created on top right l ocation
## Observation: Default (aka 1) plotting character is a circle, and the legend shows a circle next to the legend key.
## Observation: colors correspond to the same order of the legend vector. May is blue, Other is red

```


#### Scatterplot Annotation: Adding Regression Line
```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in NY City", pch = 20)) ## draw initial graph
model <- lm(Ozone ~ Wind, airquality) ## where lm is linear model
abline(model, lwd = 2)
## Observation: Regression line included 


```


### Base Graphics - Multiple base plots examples
```{r}
par(mfrow  = c(1,2))
with(airquality, 
     { 
       plot(Wind, Ozone, main = " Ozone and Wind")
       plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
     })
## Observation: par set 1 row with  2 columns  (We could have also done mfcol = c(1,2)) to achieve same result
## Observation: 2 plots, both have Ozone as Y axis

```


```{r}
# Another example with a 3rd plot
par(mfrow  = c(1,3), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
with(airquality, 
     { 
       plot(Wind, Ozone, main = " Ozone and Wind")
       plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
       plot(Temp, Ozone, main = "Ozone and Temperature")
       mtext("Ozone and Weather in NYC", outer = TRUE)
     })
## Observation: Accidently entered mfrow to be (1,2) in which it output the last plot on a seperate output.
## Observation: mtext is for master text, and it is there because we set outer marging to 0 0 2 0
```


## Base Plot Video Demonstration Follow
```{r}
# Creating random normal distributions 
x <- rnorm(100)
hist(x)
## Observation: Without actually specifying anything, default labels appear

y <- rnorm(100)
plot(x,y)
## Observation: Without actually specifying anything, default labels appear (except for title)
```


#### examples of using points
```{r}
example("points")
## Observation: will show different ways  to use points
```


#### Adding text to plots
```{r}
# Create random numbers
a <- rnorm(100)
b <- rnorm(100)
plot(a,b)

text(-2, -2, "test")
## Observation: Adds text to the indicated coordinates

legend("topleft", legend = "Data", pch = 5)
## Observation: Creats a legend on top left  with label Data using diamond symbol. 

regression <- lm(b ~ a)
abline(regression, col = "blue")
## Observation: Created linear regression onto the plot
```


```{r}

par(mfrow  = c(1,1))
## Observation: Setting up our area to view plots

x <- rnorm(100)
y <- x + rnorm(100)
## Observation: Random data  created

g <- gl(2,50)
## Observation: Group with 2 levels, 50 iterations each

g <- gl(2,50, labels = c("Male", "Female"))
## ## Observation: Same as above, but we are including labels (as shown by str(g))
str(g)


# Now we want to PLOT, here are the steps:

#1. Set  up the plotting region with no parameters
plot(x, y, type = "n")

#2. Add male group first
points(x[g == "Male"], y[g =="Male"], col = "green")

#3. Add female group
points(x[g == "Female"], y[g =="Female"], col = "blue")
## Observation: Blue circles for females, and green circles for males are  shown 
```






